Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-12-05 article.tex
Reorder

\section{\lstinline!LLVM--! instruction set}

@@

@@

@@

@@

%* list of methods for coalescing/allocating:
%   * ocamlgraph builtin
%   * greedy 
%   * briggs
%   * george
%   * welsh-powell?
\section{Graph Coloring}
The graph coloring by simplification as described by Appel

@@

@@

@@

%Runtime performance of generated code can be measured 
%Measuring the performance of a compiler backend involves assessing how well it translates high-level programming code into machine code, considering factors like execution speed and generated code size. Here are several methods to evaluate the performance of a given compiler:

@@

\begin{center}
\begin{tabular}{|c||c|c|c|c|} 
 \hline
 fib & clang & greedy & simple & linear \\ [0.5ex] 
 \hline
 40 & 1.000000 & 1.438693 & 1.441270 & 0.73 \\ [1ex] 
 41 & 1.000000 & 1.441035 & 1.440626 & 0.73 \\ [1ex] 
 42 & 	1.000000 & 1.447140 & 1.446094 & 0\\
 \hline
\end{tabular}
\end{center}

>>>>

@@

\subsection{Building a graph}

\subsection{Parameterized over individual instructions}

\subsection{Parameterized over basic blocks}

@@

\subsection{Dataflow Analysis}

@@

\subsection{Constructing an interference graph}

@@

%* list of methods for coalescing/allocating:
%   * ocamlgraph builtin
%   * greedy 
%   * briggs
%   * george
%   * welsh-powell?
\section{Graph Coloring}
The graph coloring by simplification as introduced in the Appel text takes an iterative approach.

\subsection{Coloring by simplification}

@@

Coalescing is the process of eliminating moves/copies of data from one GPR to another by combining their interference graph nodes.

@@

\subsection{\lstinline!LLVM--! instruction set}

%%%
\subsection{Translating to x86}

@@

%Runtime performance of generated code can be measured

There are different approaches to assessing how well a compiler backend translates IR to machine code targeting a specific architecture and platform. One of the primary means is to simply measure the time it takes to execute the code generated by it.

Since the target architecture is x86-64, all benchmarks are executed natively on the fastest processor available to be used consistently at the time of writing. This is to maximize the sample size and in turn reduce uncertainty, but the clock speed at which the benchmarks are performed is irrelevant to how fast the generated code is. Because of this, the metric recorded is the amount of CPU cycles spent executing from start to finish instead of actual time in seconds, as the time measure only works as a more imprecise estimate of the amount of underlying work/execution steps performed on the processor.

Needless to say, CPU cycles isn't a perfect measurement either, but it works to eliminate the clock speed as well as negate much of the time the scheduler allocates for other processes. Scheduling still has a negative impact on execution because of the cache misses caused by context switching, but outside of executing each program in immediate mode which isn't possible on Linux or macOS this is a sensible estimate. To further reduce context switching, benchmarks are made on a fresh restart with minimal background processes running.

Additionally, all benchmarks are deterministic, so the value of each virtual variable will stay the same at every instruction step of execution when compared to another run with the same starting conditions. %all the way through the execution compared to another run of the same program with the same input. %% FIXME: not necessarily true as loops exist
This means that the cycles measured over infinitely many runs will converge towards a 'perfect' run with no cache misses as caused by context switching. So across \(n\) runs, the run least affected by context switching will be the one with the least CPU cycles, so is the one most representative of the actual performance without noise.

Measurements are made with the \lstinline!perf! utility, which relies on the \textit{hardware counters} (HC) registers of modern microprocessors, which are special purpose registers meant to record performance data live during execution. Because the analysis is integrated into the chip on which it is executing, very little overhead is incurred, and is therefore the go-to for estimating native performance.

\subsubsection{Measurements}

Each \lstinline!.ll! file present in the \lstinline!benches! directory is listed here in a table denoting the input paramater on the left column and type of allocator used.

%for several reasons, one of which is to negate the clock speed as this holds no relevance to the compiler.

%considering factors like execution speed and generated code size. Here are several methods to evaluate the performance of a given compiler:

@@

\paragraph{fib.ll}
The 'dumb' fib is a very naive approach to finding the \(n\)th number of the Fibonacci sequence by calling itself recursively twice (decrementing \(n\) before each), which causes an exponential growth in function calls.
\begin{center}
\begin{tabular}{|c||c|c|c|c|} 
 \hline
 fib & clang & greedy & simple & linear \\ [0.5ex] 
 \hline
 40 & 1.000000 & 1.438693 & 1.441270 & 0.73 \\ [1ex] 
 41 & 1.000000 & 1.441035 & 1.440626 & 0.73 \\ [1ex] 
 42 & 	1.000000 & 1.447140 & 1.446094 & 0\\
 \hline
\end{tabular}
\end{center}
